Making Sequential Programs Deterministic

### 2019-04-04 design.tex
Threats to validity, and changes suggested by Jason

Both collections and map may be \<Det>, \<OrderNonDet>, or \<NonDet>, and
the element types can be specified independently of the collection type,
including qualifiers.

>>>>

Both collections and maps may be \<Det>, \<OrderNonDet>, or \<NonDet>, and
the element types can be specified independently of the collection type,
including qualifiers.

### 2019-04-04 introduction.tex
Improve a sentence based on Jason's suggestion
Threats to validity, and changes suggested by Jason

Two well-known sources of nondeterminism are concurrency
% (due to OS scheduling or message delivery over a network)
and coin-flipping
(calls to a \<random> API\@).
It may be surprising that nondeterminism is common even in sequential
programs that do not flip coins.
For example, any program that iterates over a set or map, and that uses the
results to affect the output, may produce different output on different runs.
So may any program that uses default formatting, such as Java's
\<Object.toString()> which includes an address or hash code.
Other nondeterministic APIs include date-and-time functions and
accessing system properties such as the file system or environment variables.

We have created an analysis that detects nondeterminism or verifies its
absence.
Our analysis permits programmers to specify which parts of their program
are intentionally nondeterministic, and it verifies that the remainder is deterministic.
%
Our analysis is sound:  if it issues no warnings, then the program does not
suffer nondeterminism of the given types.  (If the program uses concurrency,
it might still be nondeterministic.)
%
Our analysis works at compile time, giving a guarantee over every possible
execution of the program, by contrast to unsound dynamic tools that attempt
to discover when a program has exhibited nondeterministic behavior on
specific runs.  There is no need for a special JVM nor rerunning a program
multiple times.
%
Our analysis handles collections that will contain the same values, but
possibly in a different order, on different runs.
%
Our analysis is precise enough for practical use.  It permits calls to
nondeterministic APIs, and only issues a warning if they are used in ways
that may lead to nondeterministic output observed by a user.  Like any
sound analysis, it can issue false positive warnings.

@@

% LocalWords:  NONDEX NonDex DeFlaker Det OrderNonDet NonDet

>>>>

Two well-known sources of nondeterminism are concurrency
% (due to OS scheduling or message delivery over a network)
and coin-flipping
(calls to a \<random> API\@).
It may be surprising that nondeterminism is common even in sequential
programs that do not flip coins.
For example, a program that iterates over a hash table
may produce different output on different runs.
So may any program that uses default formatting, such as Java's
\<Object.toString()> which includes an address or hash code.
Other nondeterministic APIs include date-and-time functions and
accessing system properties such as the file system or environment variables.

We have created an analysis that detects nondeterminism or verifies its
absence.
Our analysis permits programmers to specify which parts of their program
are intentionally nondeterministic, and it verifies that the remainder is deterministic.
%
Our analysis is largely sound:  if it issues no warnings, then the program does not
suffer nondeterminism of the given types, up to the limitations of the
analysis (see \cref{sec:threats}).  If the program uses concurrency,
it might still be nondeterministic.)
%
Our analysis works at compile time, giving a guarantee over every possible
execution of the program, by contrast to unsound dynamic tools that attempt
to discover when a program has exhibited nondeterministic behavior on
specific runs.  There is no need for a special JVM nor rerunning a program
multiple times.
%
Our analysis handles collections that will contain the same values, but
possibly in a different order, on different runs.
%
Our analysis is precise enough for practical use.  It permits calls to
nondeterministic APIs, and only issues a warning if they are used in ways
that may lead to nondeterministic output observed by a user.  Like any
sound analysis, it can issue false positive warnings.

@@

\todo{It is essential that the introduction includes an example real-world
  defect that \theDeterminismChecker found.}

% LocalWords:  NonDex DeFlaker Det OrderNonDet NonDet

### 2019-04-04 java-design.tex
Threats to validity, and changes suggested by Jason

\codeid{System.getProperty()} returns a \<NonDet> type since we can't provide any determinism guarantees for
executions across machines. However, we make exceptions when the argument to this method is one of 
\codeid{"line.separator", "path.separator", or "file.separator"} as they return OS dependent values which are guaranteed to be deterministic on that OS.

>>>>

The return type of \<System.getenv}, which reads an operating system
environment variable, is always \<NonDet>.
The return type of \codeid{System.getProperty} is usually \<NonDet>. However,
the result is \<Det> if the argument is \codeid{"line.separator"}; although
it varies by operating system, many programs including \<diff> hide those
differences from the user.
The result is also deterministic for \codeid{"path.separator"} and
\codeid{"file.separator"}, because these lead to the same behavior given
corresponding environments.
The user can specify input properties that must be passed on the \<java>
command line and thus act like inputs to the program; these are also
treated as deterministic.
The inputs to a program are also treated as \<Det>.
